What is an Ancestral Recombination Graph?

### 2022-02-17 paper.tex
Try putting the shoulders on the bottom of Fig 3B.
Update

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}); the following is useful
to illustrate the computational properties of the structure.
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

%% Edges
\draw (s1) |- (s3);
\draw (s1) |- (s4) |- (s6);
\draw (s3) |- (s5);
\draw (s0) |- (s5);
\draw (s2) |- (s6);
\draw (s5) |- (s7);
\draw (s6) |- (s7);

@@

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). When we want to recover
the local tree at a position $x$ along the genome,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents; if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the first edge, and otherwise follow the second edge.
This ordering requirement, which although straightforward
to describe, has some drawbacks in practice. For example,
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
We require only two changes. Firstly, we let nodes in the
graph model \emph{genomes} rather than \emph{events}. As we
see in the next section, this allows us to model complex
patterns of inheritance without classifying different types of node.
The second change we must make is to encode
intervals of ancestry inheritance \emph{explicitly} on graph edges,
rather than \emph{implicitly} as breakpoints associated with
recombination events and edge ordering rules.
That is, given two genomes $p$
and $c$ where $p$ is the parent genome and $c$ is the child
(although many generations may have elapsed between the two),
we associate a set $I$ of intervals of genome which $c$ has
inherited from $p$.
These simple changes required to translate from the
classical Griffiths approach to the proposed encoding
are shown in Fig.\ref{fig-arg-data-structure}.

\textcolor{red}{
This falls a bit flat here,  but it's not clear how we're dividing
the responsibility for explaining the model between the
different sections.
Basically, we can't explain everything at
the end of this section, but we want to give the reader the
basic idea (which is genuinely simple). Can see how things are
shaping up after we've got a draft of the rest of the
sections, I guess.}

\section*{Genome ARGs}

The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\caption{\label{fig-pedigree-and-arg}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a (potentially) recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited. An ARG is a retrospective
structure defined with
respect to some set of sample genomes (A--D here), and so only
genomes directly ancestral to those samples are included.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination.

@@

@@

>>>>

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
The basic unit of an EARG are \emph{events} and correspond directly
to the events that
occur in the ARG stochastic process described in the previous section.
Each event corresponds to a node in the graph, and there are therefore
two types of node: common ancestor and recombination. As in the
stochastic process, edges represent lineages, and those lineages
are modified by the events that they pass through.

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

%% Edges
\draw (s1) -| (s3);
\draw (s1) -| (s4) |- (s6);
\draw (s3) |- (s5);
\draw (s0) |- (s5);
\draw (s2) |- (s6);
\draw (s5) |- (s7);
\draw (s6) |- (s7);

@@

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). When we want to recover
the local tree at a position $x$ along the genome,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents; if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the first edge, and otherwise follow the second edge.
This ordering requirement, which although straightforward
to describe, has some drawbacks in practice. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of "ARG"
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
We require only two changes. Firstly, we let nodes in the
graph model \emph{genomes} rather than \emph{events}. As we
see in the next section, this allows us to model complex
patterns of inheritance without classifying different types of node.
The second change we must make is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules.
% That is, given two genomes $p$
% and $c$ where $p$ is the parent genome and $c$ is the child
% (although many generations may have elapsed between the two),
% we associate a set $I$ of intervals of genome which $c$ has
% inherited from $p$.
These changes are straightforward, involve no loss of information
and greatly increase both the expressivity and computational
power of ancestral recombination graph data structures.

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\end{center}
\caption{\label{fig-pedigree-and-arg}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a (potentially) recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or GARG.
Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of a GARG.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-pedigree-and-arg}A
has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination. The pedigree is shown here for concreteness, but
is not a necessary part of the model. We can reason directly about the
genomes without necessarily knowing how they are combined within
individuals. This ``genome-perspective'' is show in
Fig.~\ref{fig-pedigree-and-arg}B, which depicts the same graph
of the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-pedigree-and-arg}C).

@@

% Asking the reader to flick back to figure 2 here, not great.
The differences between EARGs
and GARGs are subtle but important, and illustrated in
Fig.~\ref{fig-arg-data-structure}, which shows a simple ARG
with three samples and one recombination.
Fig.~\ref{fig-arg-data-structure}A shows this topology as a
classical Griffiths-like EARG, and
Fig.~\ref{fig-arg-data-structure}B shows it as a GARG.
The two are almost identical, except we model the recombination
event via the two parental genomes in nodes $3$ and $4$ and explicitly
store the genomic interval that the child inherited from the
parent with every edge.
This approach may seem wasteful since we must store an extra
node for each recombination and genomic interval information
with every edge (even though many simply span the full genome).
However, asymptotically the amount of information stored is the
same, and as we see in the next section, keeping interval
information for all edges enables useful generalisations.
There is also much simplicity to be gained by having
a single type of node and by treating all edges in the same way.

@@

\item We make the point about expressivity of the EARG pretty
hard, but we don't mention anything about indels and structural
rearrangements etc. It's probably worth mentioning genome graphs
somewhere just to fend off this criticism. The high-level point
would be that it's unlikely to be easier to model this stuff
using node annotations.
